Multi-state modelling with R: the {\tt msm} package \vskip 0.2in \large{Version <<echo=FALSE,results=tex>>= cat(version) @ \vskip 0.1in <<echo=FALSE,results=tex>>= cat(format(Sys.time(), " @ }

### 2014-01-18 pkg/inst/doc/msm-manual.Rnw
0.3 release, adding compiled PDF manual
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@47 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The matrix exponential $\Exp$ is different from a scalar exponential.
The exponential of a matrix is defined by the same "power series"
$\Exp(X) = 1 + X^2/2! + X^3/3! + ...$ as the scalar exponential,
except that each term $X^k$ in the series is defined by matrix products,
not element-wise scalar multiplication.  It is notoriously difficult
to calculate reliably, as discussed by Moler and van Loan \cite{matrixexp}.
For simpler models, it is feasible to calculate an analytic expression
for each element of $P(t)$ in terms of $Q$.  This is generally faster
and avoids the potential numerical instability of calculating the
matrix exponential.  Symbolic algebra sofware, such as Mathematica,
can be helpful for obtaining these expressions.  For example, the
three-state illness-death model with no recovery has a transition
intensity matrix of
\[
Q = \left(
  \begin{array}{llllll}
    -(q_{12} + q_{13}) & q_{12} & q_{13}\\
    0 & - q_{23} & q_{23}\\
    0 & 0 & 0\\
  \end{array}
\right )
\]
The corresponding time $t$ transition probabilities are
\begin{eqnarray*}
  p_{11}(t) & = & e^{-(q_{12} + q_{13})t}\\
  p_{12}(t) & = &
\left\{
\begin{array}{ll}
\frac{ q_{12}(-1 + e^{(q_{12}+q_{13}-q_{23})t})e^{-(q_{12}+q_{13})t} } { (q_{12} + q_{13} - q_{23}) } & (q_{12} + q_{13} \neq q_{23})\\
q_{12}te^{(-(q_{12} + q_{13})t} &  (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{13}(t) & = &
\left\{
\begin{array}{ll}
\frac{1 + (-q_{13} + q_{23})e^{-(q_{12}+q_{13})t}} {q_{12} + q_{13} - q_{23}} - \frac { q_{12}e^{-q_{23}t}} {q_{12} + q_{13} - q_{23} } & (q_{12} + q_{13} \neq q_{23})\\
 (-1 + e^{(q_{12} + q_{13})t} - q_{12}t)e^{-(q_{12} + q_{13})t} & (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{21}(t) & = & 0\\
  p_{22}(t) & = & e^{-q_{23}t}\\
  p_{23}(t) & = & 1 - e^{-q_{23}t}\\
  p_{31}(t) & = & 0\\
  p_{32}(t) & = & 0\\
  p_{33}(t) & = & 1\\
\end{eqnarray*}
The \Rpackage{msm} package calculates $P(t)$ analytically for selected
2, 3, 4 and 5-state models, illustrated in
Figures \ref{fig:anp2}--\ref{fig:anp5}.  For other models, which can have any transition
structure on any number of states in principle, $P(t)$ is
determined from the matrix exponential.  This is calculated using eigensystem
decomposition (if eigenvalues are distinct) or a method based on
Pad\'e approximants with scaling and squaring \cite{matrixexp} (if
there are repeated eigenvalues). Notice that the states are not
labelled in these figures.  Each graph can correspond to several
different $Q$ matrices, depending on how the states are labelled.  For
example, Figure 1 a) illustrates the model defined by either \( Q =
\left(
  \begin{array}{ll}
    - q_{12} & q_{12} \\
    0 & 0
\end{array}
\right)
\)
or
\(
Q = \left( \begin{array}{ll}
0 & 0\\
q_{21} & - q_{21}
\end{array}
\right)
\).

@@

The likelihood is maximised by numerical methods, which need a set of
initial values to start the search for the maximum.  For reassurance
that the true maximum likelihood estimates have been found, models
should be run repeatedly starting from different initial
values. However a sensible choice of initial values can be important
for unstable models with flat or multi-modal likelihoods.  For
example, the transition rates for a model with misclassification could
be initialised at the corresponding estimates for an approximating
model without misclassification. Initial values for a model without
misclassification could be set by supposing that transitions between
states take place only at the observation times.  If we observe
$n_{rs}$ transitions from state $r$ to state $s$, and a total of $n_r$
transitions from state $r$, then $q_{rs} / q_{rr}$ can be estimated by
$n_{rs} / n_r$. Then, given a total of $T_r$ years spent in state $r$,
the mean sojourn time $1 / q_{rr}$ can be estimated as $T_r / n_r$.
Thus, $n_{rs} / T_r$ is a crude estimate of $q_{rs}$. The \Rpackage{msm}
package provides a function \Rfunction{crudeinits.msm} for calculating
initial values in this way (for non-hidden Markov models).

@@

%% TODO explain about gen.inits=TRUE here

@@

A \Rfunarg{fixedpars} statement can also be useful for fixing covariate
effect parameters to zero, that is to assume no effect of a covariate
on a certain transition rate.

>>>>

The matrix exponential $\Exp$ is different from a scalar exponential.
The exponential of a matrix is defined by the same "power series"
$\Exp(X) = 1 + X^2/2! + X^3/3! + ...$ as the scalar exponential,
except that each term $X^k$ in the series is defined by matrix products,
not element-wise scalar multiplication.  It is notoriously difficult
to calculate reliably, as discussed by Moler and van Loan \cite{matrixexp}.
For simpler models, it is feasible to calculate an analytic expression
for each element of $P(t)$ in terms of $Q$.  This is generally faster
and avoids the potential numerical instability of calculating the
matrix exponential.  Symbolic algebra sofware, such as Mathematica,
can be helpful for obtaining these expressions.  For example, the
three-state illness-death model with no recovery has a transition
intensity matrix of
\[
Q = \left(
  \begin{array}{llllll}
    -(q_{12} + q_{13}) & q_{12} & q_{13}\\
    0 & - q_{23} & q_{23}\\
    0 & 0 & 0\\
  \end{array}
\right )
\]
The corresponding time $t$ transition probabilities are
\begin{eqnarray*}
  p_{11}(t) & = & e^{-(q_{12} + q_{13})t}\\
  p_{12}(t) & = &
\left\{
\begin{array}{ll}
\frac{ q_{12} }{q_{12} + q_{13} - q_{23} } (e^{-q_{23} t} - e^{-(q_{12} + q_{13})t})  & (q_{12} + q_{13} \neq q_{23})\\
q_{12}te^{(-(q_{12} + q_{13})t} &  (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{13}(t) & = &
\left\{
\begin{array}{ll}
1 - e^{-(q_{12} + q_{13})t} - \frac{ q_{12} }{q_{12} + q_{13} - q_{23} } (e^{-q_{23} t} - e^{-(q_{12} + q_{13})t}) & (q_{12} + q_{13} \neq q_{23})\\
 (-1 + e^{(q_{12} + q_{13})t} - q_{12}t)e^{-(q_{12} + q_{13})t} & (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{21}(t) & = & 0\\
  p_{22}(t) & = & e^{-q_{23}t}\\
  p_{23}(t) & = & 1 - e^{-q_{23}t}\\
  p_{31}(t) & = & 0\\
  p_{32}(t) & = & 0\\
  p_{33}(t) & = & 1\\
\end{eqnarray*}
The \Rpackage{msm} package calculates $P(t)$ analytically for selected
2, 3, 4 and 5-state models, illustrated in
Figures \ref{fig:anp2}--\ref{fig:anp5}.  For other models, which can have any transition
structure on any number of states in principle, $P(t)$ is
determined from the matrix exponential.  This is calculated using eigensystem
decomposition (if eigenvalues are distinct) or a method based on
Pad\'e approximants with scaling and squaring \cite{matrixexp} (if
there are repeated eigenvalues). Notice that the states are not
labelled in these figures.  Each graph can correspond to several
different $Q$ matrices, depending on how the states are labelled.  For
example, Figure 1 a) illustrates the model defined by either \( Q =
\left(
  \begin{array}{ll}
    - q_{12} & q_{12} \\
    0 & 0
\end{array}
\right)
\)
or
\(
Q = \left( \begin{array}{ll}
0 & 0\\
q_{21} & - q_{21}
\end{array}
\right)
\).

@@

The likelihood is maximised by numerical methods, which need a set of
initial values to start the search for the maximum.  For reassurance
that the true maximum likelihood estimates have been found, models
should be run repeatedly starting from different initial
values. However a sensible choice of initial values can be important
for unstable models with flat or multi-modal likelihoods.  For
example, the transition rates for a model with misclassification could
be initialised at the corresponding estimates for an approximating
model without misclassification. Initial values for a model without
misclassification could be set by supposing that transitions between
states take place only at the observation times.  If we observe
$n_{rs}$ transitions from state $r$ to state $s$, and a total of $n_r$
transitions from state $r$, then $q_{rs} / q_{rr}$ can be estimated by
$n_{rs} / n_r$. Then, given a total of $T_r$ years spent in state $r$,
the mean sojourn time $1 / q_{rr}$ can be estimated as $T_r / n_r$.
Thus, $n_{rs} / T_r$ is a crude estimate of $q_{rs}$.

Such default initial values can be used by supplying
\Rfunarg{gen.inits=TRUE} in the call to \Rfunction{msm} below, along
with a \Rfunarg{qmatrix} whose non-zero entries represent the allowed
transitions of the model.  Alternatively the function \Rfunction{crudeinits.msm} 
could be used to get this matrix of initial values explicitly as follows.  
These methods are only available for non-hidden Markov models.

@@

@@

A \Rfunarg{fixedpars} statement can also be used for fixing covariate
effect parameters to zero, that is to assume no effect of a covariate
on a certain transition rate.
